\section{Discussion and Future Work}\label{sec:discussion}

The calibration and scenario results validated the core claim of this study: a modified Heston process with a state-dependent $\theta$-function can produce realistic IV surfaces from structural return dynamics alone. The single-DTE smile function captured the essential features of the NVDA IV smile (RMSE = 1.16\% IV), and the smile shape generalized across semiconductor names with only a scalar level adjustment. The multi-expiration calibration (Section~\ref{sec:calibration}) extended this to the full IV surface: the five-parameter $\psi$ function with quadratic DTE term captured the U-shaped ATM term structure (RMSE = 2.31\% IV across seven expirations), confirming that the term-structure geometry could be identified from a single cross-sectional snapshot. The equilibrium initialization $v_0 = \theta(t\!=\!0)$ eliminated $v_0$ as a free parameter and produced the smile, skew, and term structure by construction.

\subsection{Limitations}

\paragraph{Term-structure calibration.} The multi-expiration calibration resolved the earlier limitation of fixing $\beta_1$, but revealed that the ATM term structure was not a simple power law---the U-shape required a quadratic extension ($\beta_5$). The fitted $\beta_1 < 0$ contradicted the initial contango assumption ($\beta_1 = 0.20$), demonstrating that the dominant short-term effect was IV \emph{decay} from gamma compression rather than contango. Whether this U-shaped structure persists across market regimes or flattens during low-volatility environments remains an open question that time-series calibration could address.

\paragraph{Cross-ticker smile shape.} The shared $(\beta_2, \beta_3, \beta_4)$ from NVDA did not perfectly transfer to all tickers---RMSE increased from 1.2\% (NVDA) to 4.7\% (INTC). The residual analysis (Figure~\ref{fig:residuals}) revealed systematic patterns in the wings, suggesting that per-ticker smile calibration would be necessary for applications requiring sub-2\% IV accuracy across all names.

\paragraph{Static calibration.} The current calibration used single-date snapshots. Time-series calibration---fitting $\theta_{s_t}$ values across multiple dates with known HMM states---would produce a more robust mapping from regime states to variance levels and would allow the mood sensitivity $\gamma$ to be identified from data rather than set heuristically.

\subsection{Future work}

\paragraph{Cross-ticker term structure.} The multi-expiration calibration used NVDA data only. Collecting cross-sectional snapshots for additional tickers would determine whether the term-structure parameters ($\beta_1$, $\beta_5$) are universal or require per-ticker fitting, analogous to the smile-shape parameters.

\paragraph{Downstream ML applications.} The primary intended use of this framework is generating realistic synthetic training data for machine-learning models in finance. Concrete applications include: training option price predictors, learning delta-hedging policies via reinforcement learning, and augmenting scarce real option data for transfer learning.

\paragraph{Greeks and risk sensitivities.} The framework naturally supports computation of Greeks ($\Delta$, $\Gamma$, $\mathcal{V}$, $\Theta$) via finite differences on the CRR tree. Including these in the synthetic dataset would support hedging applications.

\paragraph{Cross-asset IV correlation.} The current mood mechanism created correlated IV dynamics through a shared stress signal. A richer model could introduce ticker-pair IV correlations through the copula structure, producing a full cross-asset IV surface that respects observed correlation patterns between individual-name and index volatility.
